---
title: "VRBIS Data Cleaning"
author: "Beth Jump"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## set up the environment
```{r }
# check to see the default library that's set
.libPaths()

# can select the primary libpath using this code. Replace the [2] with the number libpath you want to use
.libPaths(.libPaths()[2])

# install and save these packages:
install.packages(c("icd", "stringr", "dplyr", "sas7bdat", "eeptools"))
library("icd") 
library("stringr") 
library("dplyr")
library("sas7bdat")
library("eeptools")

#disable scientific notation so the SFN & LFN's are read in as numbers not in scientific notation
options(scipen = 999)
```

## set up file paths you'll need:
```{r}
# file paths for: location of county VRBIS data, location of reallocates VRBIS data, location to save cleaned resident data set, location to save cleaned non-resident data set
County_path <- ""
reallocates_path <- ""
resident_clean <- ""
nonres_clean <- ""
```

## import csv files for deaths from VRBIS into R
```{r import file}
county_deaths <- read.csv(County_path, header = F)
Reall_deaths <- read.csv(reallocates_path, header = F)
```

## add column names to the datasets 
```{r cleaning}
# create a character vector with labels of columns
var1 <- c("SFN_1", "LRN_2", "fname_3", "mname_4", "lname_5", "recstatus_6", "trunccode_7", "dob_8","byyyy_9", "bmm_10", "bdd_11", "ageyears_12", "agemons_13", "agedays_14", "agehours_15", "agemins_16", "agetype_17", "ageunit_18", "sex_19", "dod_20", "dhour_21", "fnd_22", "est_23", "dyyyy_24", "dmm_25", "ddd_26","birthstateforeigntext_27", "birthstateforeigncode_28", "birthcountrytext_29", "birthstateprovince_30", "ssn_31", "armedforces_32", "maritalstatus_33", "educlevelcacode_34", "educlevelnchscode_35", "hispanic_36", "hispanictypecode_37", "hispanicspecifytext_38", "race1code_39", "race2code_40", "race3code_41", "race1text_42", "race2text_43", "race3text_44", "race1codefinal_45", "race2codefinal_46", "race3codefinal_47", "race1textnchs_48", "race2textnchs_49", "race3textnchs_50", "multirace_51", "occupation_52", "industry_53", "yearsinocc_54", "resstreetnum_55", "resstreetname_56", "rescitytext_57", "resccode_58", "rescountycdphcode_59", "rescountytext_60", "rescountynchscode_61", "rescountyfinalcode_62", "rescountyfinaltext_63", "rescountycdphcode2_64", "rescountynchscode2_65", "reszipcode5_66", "reszipcode4_67", "stateForeignCountryText_68", "stateForeignCountrycode_69", "rescountrycode_70", "resstaeprovince_71", "yrsincounty_72", "rescensustract_73", "reslatitude_74", "reslongitude_75", "informantrel_76", "informantfname_77", "informantmname_78", "informantlname_79", "informantmailstreetnum_80", "informantstreetname_81", "informantapt_82", "informantcity_83", "informantstate_84", "informantzip5_85", "informantzip4_86", "survspousefname_87", "survspousemname_88", "survspouselname_89", "fatherfname_90", "fathermname_91", "fatherlname_92", "fatherBirthStateText_93", "fatherBirthStateCode_94", "fatherBirthCountryCode_95", "fatherBirthCountryCodeFinal_96", "motherfname_97", "mothermname_98", "motherlname_99", "motherBirthStateText_100", "motherBirthStateCode_101", "motherBirthCountryCode_102", "fatherBirthCountryCodeFinal_103", "dispDate_104", "dispPlacedesc_105", "dispPlaceStreetNum_106", "dispPlaceStreetName_107", "dispPlaceStreetCityCode_108", "dispPlaceStreetState_109", "dispPlaceStreetStateCode_110", "dispPlaceStreetCountryCode_111", "dispPlaceStreetZip5_112", "dispPlaceStreetZip4_113", "dispTypeText_114", "dispTypeCode_115", "embalmerSig_116", "embalmerLic_117", "funeralHomeName_118", "funeralHomeLic_119", "localRegstrarName_120", "localRegDate_121", "deathFacilityNameLocation_122", "deathInHospCode_123", "deathOutHospCode_124", "placeOfDeathFacility_125", "deathCountyText_126", "deathCountyCode_127", "deathCountyCode2_128", "LocalRegDistrictCode_129", "deathState_130", "deathCountry_131", "placeDeathStreetNum_132", "placeDeathStreetName_133", "cityofPlaceDeath_134", "causeOfDeath_135", "intervalOnsetDeath_136", "causeOfDeath1_137", "intervalOnsetDeath1_138", "causeOfDeath2_139", "intervalOnsetDeath2_140", "causeOfDeath3_141", "intervalOnsetDeath3_142", "prelimCode_143", "finalICDCode_144", "finalICDCodeDate_145", "Coroner_146", "coronerRefNum_147", "biopsy_148", "autopsy_149", "autopsyUsed_150", "otherSigCond_151", "operation_152", "operationCode_153", "pregStatus_154", "datePhyAtt_155", "datePhySawAlive_156", "certifierSig_157", "certifierNameTitle_158", "certifierLic_159", "certifierSigDate_160", "attPhyName_161", "attPhyStreetNum_162", "attPhyStreetName_163", "attPhyCity_164", "attPhyState_165", "attPhyZip5_166", "attPhyZip4_167", "mannerOfDeath_168", "injuredAtWork_169", "injuryDate_170", "injuryYear_171", "injuryMonth_172", "injuryDay_173", "injuryHour_174", "placeInjuryText_175", "placeInjuryCode_176", "envirPlaceInjuryCode_177", "injuryDesc_178", "injPLaceStreetNum_179", "injPLaceStreetName_180", "injPLaceCity_181", "injPLaceState_182", "injPLaceZip5_183", "injPLaceZip4_184", "coronerSig_185", "coronerSigDate_186", "coronerNameTitle_187","certifierType_188", "stateRegDate_189", "monthProcessed_190", "exportDate_191", "rescensustract_192", "reslatitude_193", "reslongitude_194", "deathcensustract_195", "deathlatitude_196", "deathlongitude_197", "injurycensustract_198", "injurylatitude_199")

var2 <- c("injurylongitude_200", "ea1", "ea2", "ea3", "ea4", "ea5", "ea6", "ea7", "ea8", "ea9", "ea10", "ea11", "ea12", "ea13", "ea14", "ea15", "ea16", "ea17", "ea18", "ea19", "ea20", "ra1", "ra2", "ra3", "ra4", "ra5", "ra6", "ra7", "ra8", "ra9", "ra10", "ra11", "ra12", "ra13", "ra14", "ra15", "ra16", "ra17", "ra18", "ra19", "ra20", "exportdate")

vars <- c(var1, var2)

# add the column names to each dataset
colnames(county_deaths) <- vars
colnames(Reall_deaths) <- vars
```

## split out the data into two datasets: residents who died in the county and non-residents who died in the county
```{r}
# only keep deaths for county residents (ex: San Mateo County is county code 41)
county_deaths_only <- county_deaths[which (county_deaths$rescountycdphcode_59 == "41"),]

# create separate dataset for non-residents who died here
nonres <- county_deaths[(county_deaths$rescountycdphcode_59 != "41" & county_deaths$deathCountyCode_127 == 41),]

# check distribution of reallocation deaths to make sure all have the correct county code 
table(county_deaths_only$rescountycdphcode_59)
table(Reall_deaths$rescountycdphcode_59)
```

## pull out variables for top death analysis - keep columns:
  1. state file number
  2. local reg number
  3. first name
  4. middle name
  5. last name
  8. DOB
  12. Age in years
  19. Sex
  20. Date of death
  27. Birth state/foreign country
  34. Highest education level
  45-47. Race/ethnicity codes
  51. Multi-race status
  55-57. Address
  64. County code (41)
  66. zip code
  135. primary cause of death (text)
  137. primary cause of death (text)
  139. primary cause of death 2 (text)
  141. primary cause of death 3 (text)
  143. preliminary cause of death
  144. final case of death (ICD 10 code)
  151. Other signifcant conditions
  168. Manner of death
  176. Place of injury
  221-231. multiple cause of death codes 

```{r  subset and combine 2 sets}
# subset both datasets so they only have the columns listed above
county_deaths_only1 <- county_deaths_only[,c(1:5,8,12,19,20,27,34,45:47,51,55:57,64,66,135,137,139,141,143,144,151,168,176,221:231)]
Reall_deaths1 <- Reall_deaths[,c(1:5,8,12,19,20,27,34,45:47,51,55:57,64, 66,135,137,139,141,143,144,151,168,176,221:231)]
nonres1 <- nonres[,c(1:5,8,12,19,20,27,34,45:47,51,55:57,64, 66,135,137,139,141,143,144,151,168,176,221:231)]

# convert zip code from factor to numeric in the SMC dataset so that we can combine the datasets without errors 
Reall_deaths1$reszipcode5_66 <-as.factor(Reall_deaths1$reszipcode5_66)

# add the two dataframes together to get one dataframe with all deaths (local and reallocates)
comb_deaths <- rbind(county_deaths_only1, Reall_deaths1, stringsAsFactors = FALSE)

# confirm no duplicates
dups <- duplicated(comb_deaths)
table(dups)
```

## add a column for death month
```{r}
# for county resident deaths
comb_deaths$dod_test <- as.Date(comb_deaths$dod_20, format = "%m/%d/%Y")
comb_deaths$deathmo <- as.numeric(substring(comb_deaths$dod_test,6,7))
```

## split the ICD10 codes out into string and numeric columns so we can match on strings and use the numbers to capture a range of codes
```{r}
# make sure codes are in the short, not decimal format
comb_deaths$finalICDCode_144 <- as.character(comb_deaths$finalICDCode_144)

# split ICD code into character & numeric columns
comb_deaths$icd_char <- as.character(substring(comb_deaths$finalICDCode_144,1,1))
comb_deaths$icd_num <- as.numeric(substring(comb_deaths$finalICDCode_144,2))

# make sure codes are in the short, not decimal format
nonres1$finalICDCode_144 <- as.character(nonres1$finalICDCode_144)

# split ICD code into character & numeric columns
nonres1$icd_char <- as.character(substring(nonres1$finalICDCode_144,1,1))
nonres1$icd_num <- as.numeric(substring(nonres1$finalICDCode_144,2))
```

## write to file
```{r}
write.csv(comb_deaths, resident_clean)
write.csv(nonres1, nonres_clean)
```
